---
title: "Using Tidycensus to Obtain Census Data"
output:
  rmdformats::html_docco:
    self_contained: true # Other options are downcute, material, readthedown, html_clean, html_docco, lockdown, https://github.com/juba/rmdformats
    thumbnails: false
    lightbox: true
    gallery: false
    highlight: kate
    number_sections: false
---

## Libraries

```{r setup, include = F}
knitr::opts_chunk$set(echo = F, cache = T, warning = F, message = F)
options(tigris_use_cache = TRUE) # For caching maps
library(tidyverse)
library(tidycensus)
library(leaflet)
library(sf)
library(viridis)
# remotes::install_github("tylermorganwall/rayshader")
# library(rayshader)
library(rgl)
library(devout)
library(devoutrgl)
library(ggrgl)
theme_set(theme_void() + theme(plot.title = element_text(hjust = 0.5)))
```

## Setup

```{r}
# Installing census api key, install only necessary once for adding to Renvironment
# census_api_key("aa7dad57625fd3a7f42a8066e1f3a2ea3ff31ed7", install = T)
# Renvironment needs to be reread
# readRenviron("~/.Renviron")
# Check key install with
# Sys.getenv("CENSUS_API_KEY")
```

By default, specifying "population", "components", or "housing" as the product in get_estimates() returns all variables associated with that component. For example, we can request all components of change variables for US states in 2017:

```{r}
us_population <- get_estimates(geography = "state", product = "population")
unique(us_population$variable)

us_components <- get_estimates(geography = "state", product = "components")
us_components

```

## A Density Example

US is CRAZY low density.

```{r}
us_county_density <- get_estimates(geography = "county",
                                   variables = "DENSITY",
                                   geometry = T,
                                   shift_geo = T)

ggplot(us_county_density, aes(fill = value, color = value)) +
  geom_sf() +
  scale_fill_distiller(palette = "Purples", direction = 1) +
  scale_color_distiller(palette = "Purples", direction = 1)
```

Take a look in 3d!

```{r}
us_state_density <- get_estimates(geography = "state",
                                   variables = "DENSITY",
                                   geometry = T,
                                   shift_geo = T)

states_map <- map_data("state") %>%
  mutate(region = str_to_title(region))
threed_map <- left_join(states_map, us_state_density, by = c("region" = "NAME"))

p <- ggplot(threed_map, aes(long, lat, group = group)) +
  geom_polygon_z(aes(fill = value, z = value), color = "white", extrude = TRUE)+
  scale_fill_viridis_c(option = "C") +
  theme_ggrgl() + 
  labs(title = "US Population Density") + 
  coord_fixed(1.2)

devoutrgl::rgldev(fov = 30, view_angle = -30)
p
invisible(dev.off())
```


## A Migration Example

```{r}
net_migration <- get_estimates(geography = "county",
                               variables = "RNETMIG",
                               geometry = TRUE,
                               shift_geo = TRUE)
order = c("-15 and below", "-15 to -5", "-5 to +5", "+5 to +15", "+15 and up")

net_migration <- net_migration %>%
  mutate(groups = case_when(
    value > 15 ~ "+15 and up",
    value > 5 ~ "+5 to +15",
    value > -5 ~ "-5 to +5",
    value > -15 ~ "-15 to -5",
    TRUE ~ "-15 and below"
  )) %>%
  mutate(groups = factor(groups, levels = order))

ggplot() +
  geom_sf(data = net_migration, aes(fill = groups, color = groups), lwd = 0.1) +
  geom_sf(data = tidycensus::state_laea, fill = NA, color = "black", lwd = 0.1) +
  scale_fill_brewer(palette = "PuOr", direction = -1) +
  scale_color_brewer(palette = "PuOr", direction = -1, guide = FALSE) +
  coord_sf(datum = NA) +
  theme_minimal(base_family = "Roboto") +
  labs(title = "Net migration per 1000 residents by county",
       subtitle = "US Census Bureau 2019 Population Estimates",
       fill = "Rate")
```



## A Couple Mapping Examples

Getting Oregon population data by county, and then getting a color for percentile using n = 10 to plot a leaflet.

```{r}
oregon_pop <- get_acs(geography = "tract",
                     variables = "B01003_001",
                     state = "OR",
                     geometry = TRUE)

pal <- colorQuantile("Blues", domain = oregon_pop$estimate, n = 10)

oregon_pop %>% 
  st_transform(crs = "+init=epsg:4326") %>%
  leaflet(width = "100%") %>%
  addProviderTiles(provider = "CartoDB.Positron") %>%
  addPolygons(popup = ~ str_extract(NAME, "^([^,]*)"),
              stroke = FALSE,
              smoothFactor = 0,
              fillOpacity = 0.7,
              color = ~ pal(estimate)) %>%
  addLegend("bottomright", 
            pal = pal, 
            values = ~ estimate,
            title = "Population percentiles",
            opacity = 1)
```

Plot a specific county by tract with ggplot.

```{r}
benton <- get_acs(geography = "tract", variables = "B19013_001",
                state = "OR", county = "Benton", geometry = TRUE)

ggplot(data = benton, aes(fill = estimate, color = estimate)) +
  geom_sf() +
  coord_sf(crs = 26914) +
  scale_fill_viridis(option = "magma") +
  scale_color_viridis(option = "magma")
```

Plot the US by county with leaflet

```{r}
us <- unique(fips_codes$state)[1:51]

totalpop_county_sf <- map_df(us, function(x) {
  get_acs(geography = "county", variables = "B01003_001",
          state = x, geometry = TRUE) # Do without geometry = T for more just population information
})

pal <- colorNumeric(palette = "viridis", domain = totalpop_county_sf$estimate) # Numeric color palette

totalpop_county_sf %>% # to filter by state do filter(grepl("Oregon", NAME)) %>%
  st_transform(crs = "+init=epsg:4326") %>%
  leaflet(width = "100%") %>%
  addProviderTiles(provider = "CartoDB.Positron") %>%
  addPolygons(popup = ~ str_extract(NAME, "^([^,]*)"),
              stroke = FALSE,
              smoothFactor = 0,
              fillOpacity = 0.7,
              color = ~ pal(estimate)) %>%
  addLegend("bottomright", 
            pal = pal, 
            values = ~ estimate,
            title = "Population Estimate",
            labFormat = labelFormat(),
            opacity = 1)
```

Ggplot the US by state excluding Alaska and the US. Also includes geom_sf_text, which amusingly labels Michigan in the Upper Peninsula.

```{r}
totalpop_state_sf <- map_df(us, function(x) {
  get_acs(geography = "state", variables = "B01003_001",
          state = x, geometry = TRUE) # Do without geometry = T for more just population information
})
# Code to do this the old way, creating an X and Y for the center of each state
# states <- st_as_sf(maps::map("state", plot = FALSE, fill = TRUE)) %>%
#   mutate(ID = str_to_title(ID)) %>%
#   rename(NAME = ID) 
# states <- states %>% 
#   bind_cols(centroid = st_geometry(st_centroid(states))) %>%
#   mutate(centroid = str_remove(centroid, "c\\(")) %>%
#   mutate(centroid = str_remove(centroid, "\\)")) %>%
#   separate(centroid, c("X", "Y"), sep = ", ") %>%
#   st_set_geometry(NULL)
# 
# totalpop_state_sf <- totalpop_state_sf %>% left_join(states, by = c("NAME"))

uspop_plot <- totalpop_state_sf %>%
  dplyr::filter(NAME != "Alaska" & NAME != "Hawaii") %>%
  ggplot(aes(fill = estimate)) +
  geom_sf() +
  geom_sf_text(aes(label = format(estimate, big.mark = ",", scientific = F)), size = 2) +
  coord_sf(crs = 26914) +
  scale_fill_distiller(palette = "Blues", direction = 1, labels = scales::comma) +
  labs(fill = "Population", title = "State Populations")
```

Function for tract by state leaflet map of home values, state parameter should be "OR" format

```{r echo = F}
stateHome <- function(state) {
  stateValue <- get_acs(geography = "tract", 
                    variables = "B25077_001", 
                    state = state,
                    geometry = TRUE)

  pal <- colorNumeric(palette = "Blues", 
                    domain = stateValue$estimate)

  stateValue %>%
    st_transform(crs = "+init=epsg:4326") %>%
    leaflet(width = "100%") %>%
    addProviderTiles(provider = "CartoDB.Positron") %>%
    addPolygons(popup = ~ str_extract(NAME, "^([^,]*)"),
                stroke = FALSE,
                smoothFactor = 0,
                fillOpacity = 0.7,
                color = ~ pal(estimate)) %>%
    addLegend("bottomright", 
              pal = pal, 
              values = ~ estimate,
              title = "Median Home Value",
              labFormat = labelFormat("$"),
              opacity = 1)
}
stateHome("CA") #Pretty sick
```

State by state leaflet of rent cost.

```{r echo = F}
stateRent <- function(state) {
  rent <- get_acs(geography = "tract", variables = "DP04_0134", 
                state = state, geometry = TRUE)
  pal <- colorNumeric(palette = "viridis", domain = rent$estimate)
  rent %>%
    st_transform(crs = "+init=epsg:4326") %>%
    leaflet(width = "100%") %>%
    addProviderTiles(provider = "CartoDB.Positron") %>%
    addPolygons(popup = ~ str_extract(NAME, "^([^,]*)"),
                stroke = FALSE,
                smoothFactor = 0,
                fillOpacity = 0.7,
                color = ~ pal(estimate)) %>%
    addLegend("bottomright", 
              pal = pal, 
              values = ~ estimate,
              title = paste("Rent in", state),
              labFormat = labelFormat("$"),
              opacity = 1)
}

stateRent(state = c("WA", "OR", "CA"))
```


Tract by state income.

```{r echo = F}
stateIncome <- function(state) {
  income <- get_acs(geography = "tract", variables = c(medincome = "B19013_001"), 
                  state = state, geometry = TRUE)
  pal <- colorNumeric(palette = "Reds", domain = income$estimate)
  income %>%
    st_transform(crs = "+init=epsg:4326") %>%
    leaflet(width = "100%") %>%
    addProviderTiles(provider = "CartoDB.Positron") %>%
    addPolygons(popup = ~ str_extract(NAME, "^([^,]*)"),
                stroke = FALSE,
                smoothFactor = 0,
                fillOpacity = 0.7,
                color = ~ pal(estimate)) %>%
    addLegend("bottomright", 
              pal = pal, 
              values = ~ estimate,
              title = paste("Income in", state),
              labFormat = labelFormat("$"),
              opacity = 1)
}

stateIncome("OR")
stateIncome("NY")
```












